---
title: "Mapping Assignment"
author: "Ruxin Liu"
date: "10/30/2020"
output: pdf_document
---

# Set up:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align = "center")
```

```{r, message=FALSE}
library(tidyverse)
library(drat)
library(hurricaneexposuredata)
addRepo("geanders")
data("hurr_tracks")
data("rain")
```

## For the storm Floyd-1999
```{r}
# Create a sub-dataset for the storm Floyd-1999 with hurricane data
Floyd_hurr <- hurr_tracks %>% 
  filter(storm_id == "Floyd-1999")
```

```{r}
# Create a sub-dataset for the storm Floyd-1999 with rain data
Floyd_rain <- rain %>% 
  filter(storm_id == "Floyd-1999")
```


```{r, message=FALSE}
# head(Floyd_hurr)
# head(Floyd_rain)
library(maps)
# Since the rain data contains fips only, needs to convert this information into
# longitude and latitude information 
# head(county.fips)
# The county.fips data contains information for fips, but fips are integers
Floyd_rain$fips <- as.integer(Floyd_rain$fips)
# head(Floyd_rain)
# Now, the fips in both tables are in a consistent format. 
```

```{r, message=FALSE}
# For each fip, there were several observations due to time lag, therefore the 
# sum will be calculated for the purpose of plotting. 
Floyd_rain <- Floyd_rain %>% 
  group_by(fips) %>% 
  summarize(precip = sum(precip), precip_max = sum(precip_max))
```

```{r, message=FALSE}
# Add the county information into the rain data set
Floyd_rain <- left_join(Floyd_rain, county.fips, by = "fips")
# A new row polyname is added at the end, needs to separate into two columns
library(magrittr)
Floyd_rain %<>% separate(polyname, c("county1", "county2"), sep = ",")
# The Floyd_rain now contains the county information as well
```

```{r}
# Check all counties showed in the rain data
# unique(Floyd_rain$county1)
counties_f <- c("alabama", "arkansas", "connecticut", "delaware", "district of columbia",
                "florida","georgia", "illinois", "indiana","iowa", "kansas", "kentucky", 
                "louisiana","maine", "maryland","massachusetts", "michigan", "mississippi",
                "missouri","new hampshire", "new jersey", "new york", "north carolina", 
                "ohio", "oklahoma", "pennsylvania", "rhode island", "south carolina",
                "tennessee", "texas","vermont","virginia", "west virginia", "wisconsin")  
# The map_data function provides longitude and latitude information for conuties
# head(map_data("county"))
counties_floyd <- map_data("county", counties_f)
state_floyd <- map_data("state", counties_f)
```

```{r}
# Rename the rain data set for consistency
Floyd_rain <-Floyd_rain %>% 
  rename(region = county1, subregion = county2)
# Add the longitude and latitude information into the rain data set
Floyd_rain <- left_join(counties_floyd, Floyd_rain,
                          by = c("region", "subregion"))
```


```{r}
# Create different range groups that will be showed on the map based on the 
# precip level
Floyd_rain <- Floyd_rain %>% 
  mutate(`Rainfall (mm)` = cut(Floyd_rain$precip, 
                     breaks = c(0, 25, 50, 75, 100, 125, 150, 175, 200, 225),
                     include.lowest = TRUE))
Floyd_rain <- na.omit(Floyd_rain)
```


## For Allison-2001:

```{r}
# Similar steps for Allison-2001
Allison_hurr <- hurr_tracks %>% 
  filter(storm_id == "Allison-2001")
Allison_rain <- rain %>% 
  filter(storm_id == "Allison-2001")
```

```{r, message=FALSE}
Allison_rain$fips <- as.integer(Allison_rain$fips)
Allison_rain <- Allison_rain %>% 
  group_by(fips) %>% 
  summarize(precip = sum(precip), precip_max = sum(precip_max))
Allison_rain <- left_join(Allison_rain, county.fips, by = "fips")
Allison_rain %<>% separate(polyname, c("county1", "county2"), sep = ",")
```

```{r}
#unique(Allison_rain$county1)
counties_a <- c("alabama", "arkansas", "connecticut", "delaware", "district of columbia",
                "florida","georgia", "illinois", "indiana","iowa", "kansas", "kentucky", 
                "louisiana","maine", "maryland","massachusetts", "michigan", "mississippi",
                "missouri","new hampshire", "new jersey", "new york", "north carolina", 
                "ohio", "oklahoma", "pennsylvania", "rhode island", "south carolina",
                "tennessee", "texas","vermont","virginia", "west virginia", "wisconsin")  
counties_allison <- map_data("county", counties_a)
state_allison <- map_data("state", counties_a)
```

```{r}
Allison_rain <-Allison_rain %>% 
  rename(region = county1, subregion = county2)
Allison_rain <- left_join(counties_allison, Allison_rain,
                          by = c("region", "subregion"))
```

```{r}
# Create 2 different groups that will be showed on the map based on the precip level
Allison_rain <- Allison_rain %>% 
  mutate(`Rain > 175 mm` = ifelse(Allison_rain$precip < 175, "Unexposed", "Exposed"))
Allison_rain <- na.omit(Allison_rain)
```


# Floyd-1999 map with ggplot2:

```{r}
ggplot() + geom_polygon(data = Floyd_rain, aes(x = long, y = lat, group = group, 
                                                 fill = `Rainfall (mm)`), 
                        color = "grey", size = 0.2, alpha = 1.6) + 
  geom_polygon(data = state_floyd, aes(x = long, y = lat, group = group),
                 color = "black", fill = "white", size = 0.2, alpha = 0.3) +
  geom_path(aes(x = Floyd_hurr$longitude, y = Floyd_hurr$latitude), color = "red") +
      scale_fill_brewer(palette = "Blues") +
  ggtitle("Floyd-1999") +
  # Center the title
  theme(plot.title = element_text(hjust = 0.5))
```



# Allison-2001 with ggplot2:

```{r}
  ggplot() + geom_polygon(data = Allison_rain, aes(x = long, y = lat, group = group, 
                                                 fill = `Rain > 175 mm`), 
                          color = "grey", size = 0.2, alpha = 1.6) +
  geom_polygon(data = state_allison, aes(x = long, y = lat, group = group),
                 color="black", fill="white", size = 0.2, alpha = 0.1) +
  geom_path(aes(x = Allison_hurr$longitude, y = Allison_hurr$latitude), color = "red") +
  # Change color manually 
  scale_fill_manual(values = c("darkblue", "white"))+ 
  ggtitle("Allison-1999") +
  theme(plot.title = element_text(hjust = 0.5))
```

# Floyd-1999 map with tmap:

```{r, message=FALSE}
library(sp)
library(tmap)
```


```{r}
library(sf)
# Since tmap requires to input sf object, need to convert foreign object to an sf object
# Obtain spatial information of counties
Floyd_sf <- st_as_sf(map('county', plot = F, fill = T))
# head(county.fips)
# It is found that the county.fips contain same county information, but with a
# different name -- change the column name in order to merge
colnames(county.fips)[2] <- colnames(Floyd_sf)[1] 
# Adding the spatial data into the rain data by merging
Floyd_sf <- left_join(Floyd_sf, county.fips, by = "ID")
Floyd_sf <- right_join(Floyd_sf, Floyd_rain, by = "fips")
``` 

```{r}
# Make the hurricane track data into the spatial line
Floyd_line <- SpatialLines(list(Lines(Line(cbind(Floyd_hurr$longitude,Floyd_hurr$latitude)),
                                      ID = "Floyd-1999")))
# Although the Floyd_hurr only contains information of Floyd-1999, but this 
# function needs to specify the ID
```

```{r, message=FALSE,warning=FALSE}
tm_shape(Floyd_sf) +
  tm_polygons('precip', title = "Rainfall (mm)", palette = "Blues", 
              breaks = c(0, 25, 50, 75, 100, 125, 150, 175, 200, 225)) +
  tm_layout(main.title = "Floyd-1999",
            main.title.position = "center") +
tm_shape(Floyd_line) +
  tm_lines(col = "red", lwd = 1.5)
```

# Allison-2001 with tmap:

```{r, message=FALSE, warning=FALSE}
# Similar steps for Allison-2001
Allison_sf <- st_as_sf(map('county', plot = F, fill = T))
colnames(county.fips)[2] <- colnames(Allison_sf)[1] 
Allison_sf <- left_join(Allison_sf, county.fips, by = "ID")
Allison_sf <- right_join(Allison_sf, Allison_rain, by = "fips")
colnames(Allison_sf)[11] <- "expose"
# Make the hurricane track data into the spatial line
Allison_line <- SpatialLines(list(Lines(Line(cbind(Allison_hurr$longitude, Allison_hurr$latitude)),
                                      ID = "Allison-2001")))
```

```{r, message=FALSE}
# Create palette for the tmap
my_pal <- c('#070385', '#FAF8F7')
tm_shape(Allison_sf)+
  tm_polygons("expose", title = "Rain > 175 mm", palette = my_pal)+
  tm_layout(main.title = "Allison-2001",
            main.title.position = "center") +
tm_shape(Allison_line)+
  tm_lines(col = "red", lwd = 1.5)
```

# Reference:
1. Help and inspiration from other teams/classmates + MA615 class notes
2. Stack Overflow: https://stackoverflow.com/questions/27998952/how-to-create-spatialline-object
3. Stack Overflow: https://stackoverflow.com/questions/58645771/how-do-i-customize-my-color-range-for-my-tmap-plot-in-r
4. http://rstudio-pubs-static.s3.amazonaws.com/5312_98fc1aba2d5740dd849a5ab797cc2c8d.html
5. https://htmlcolorcodes.com/
6. Stack Overflow: https://stackoverflow.com/questions/46550752/create-a-colour-palette-for-tm-fill
7. Tennekes M (2018). “tmap: Thematic Maps in R.” _Journal of Statistical
Software_, *84*(6), 1-39. doi: 10.18637/jss.v084.i06 (URL:
https://doi.org/10.18637/jss.v084.i06).
8. Wickham et al., (2019). Welcome to the tidyverse. Journal of Open
  Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686
9. Original S code by Richard A. Becker, Allan R. Wilks. R version by
  Ray Brownrigg. Enhancements by Thomas P Minka and Alex Deckmyn.
  (2018). maps: Draw Geographical Maps. R package version 3.3.0.
  https://CRAN.R-project.org/package=maps
10. Anderson B, Schumacher A, Crosson W, Al-Hamdan M, Yan M, Ferreri J,
Chen Z, Quiring S, Guikema S (2020). _hurricaneexposuredata: Data
Characterizing Exposure to Hurricanes in United States Counties_. R
package version 0.1.0, <URL:
https://github.com/geanders/hurricaneexposuredata>.
11.  Stefan Milton Bache and Hadley Wickham (2014). magrittr: A
  Forward-Pipe Operator for R. R package version 1.5.
  https://CRAN.R-project.org/package=magrittr
12. Dirk Eddelbuettel with contributions by Carl Boettiger, Neal Fultz,
  Sebastian Gibb, Colin Gillespie, Jan Górecki, Matt Jones, Thomas
  Leeper, Steven Pav, Jan Schulz, Christoph Stepper, Felix G.M. Ernst
  and Patrick Schratz. (2020). drat: 'Drat' R Archive Template. R
  package version 0.1.8. https://CRAN.R-project.org/package=drat
13. RDocumentation https://www.rdocumentation.org/packages/base/versions/3.6.2/topics/cut